---
title: "Class 10. Rosling assembly, 2007"
format: html
---

**Question for today:** What is the link between income and life expectancy in 2007?

```{r}
#| label: setup
pacman::p_load(tidyverse, here)
```

Rosling shows four variables at once: x is income, y is life expectancy, size is population, color is continent.

## 1. Start: two numbers

```{r}
#| label: step-1
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp)) +
  geom_point()
```

## 2. Log scale on x

Income spans 278 to 49,357 dollars. A linear axis squeezes most countries to the left. A log axis gives each doubling the same width.

```{r}
#| label: step-2
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp)) +
  geom_point() +
  scale_x_log10()
```

## 3. Size by population

`scale_size_area()` makes the area of each circle proportional to population, which is honest. The default radius scale exaggerates differences.

```{r}
#| label: step-3
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp, size = population)) +
  geom_point() +
  scale_x_log10() +
  scale_size_area(max_size = 15)
```

## 4. Color by continent

```{r}
#| label: step-4
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp, size = population, color = continent)) +
  geom_point() +
  scale_x_log10() +
  scale_size_area(max_size = 15)
```

> Deliverable: the static Rosling chart for 2007 with four variables.
