---
title: "Class 3. Recap, and the type of variable decides the chart"
format: html
---

**Question for today:** What does each variable type allow us to draw?

```{r}
#| label: setup
pacman::p_load(tidyverse, palmerpenguins, here)
```

## 0. Recap: bad graphs and the five principles

**Ask. Show. Cut. Focus. Tell.** Today we practice **Show**: honest data, the right chart for the data type, and honest scales.

Three mistakes from class 1, rebuilt with penguins.

```{r}
#| label: recap-baseline
# Mistake 1: omitting the baseline. The axis starts at 3,600 g.
# Chinstrap looks 30% heavier than Adelie. The real gap is 0.9%.
penguins |>
  group_by(species) |>
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE)) |>
  ggplot(aes(x = species, y = mean_mass)) +
  geom_col() +
  coord_cartesian(ylim = c(3600, 5100))
```

```{r}
#| label: recap-baseline-fix
# Fix: bars start at zero
penguins |>
  group_by(species) |>
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE)) |>
  ggplot(aes(x = species, y = mean_mass)) +
  geom_col()
```

```{r}
#| label: recap-wrong-graph
# Mistake 2: using the wrong graph. A line joins categories that have no order.
penguins |>
  group_by(species) |>
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE)) |>
  ggplot(aes(x = species, y = mean_mass, group = 1)) +
  geom_line() +
  geom_point()
```

```{r}
#| label: recap-axis
# Mistake 3: manipulating the y axis. Same data, a range of 0 to 20,000 g.
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  coord_cartesian(ylim = c(0, 20000))
```

Which principle failed in each graph? Write it down. The rest of today answers: **which chart fits which type of data?**

## 1. Know the data before you plot

```{r}
#| label: glimpse
# dbl and int are numbers, chr and fct are categories, date is time
penguins |>
  glimpse()
```

```{r}
#| label: missing
# Count missing values in each column
penguins |>
  summarize(across(everything(), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

## 2. One number: histogram

```{r}
#| label: histogram
penguins |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250)
```

## 3. Two numbers: scatter plot

```{r}
#| label: scatter
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

## 3b. Check for outliers

A histogram and a scatter plot show values that sit far from the rest. An outlier needs a check before you remove it.

```{r}
#| label: outliers-histogram
# Which penguins are in the right tail of the histogram?
penguins |>
  slice_max(body_mass_g, n = 5) |>
  select(species, island, sex, body_mass_g, flipper_length_mm)
```

```{r}
#| label: outliers-scatter
# Color by species. A point that looks extreme overall may be typical for its group.
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

For each extreme value, write one sentence: **data error** (impossible value, wrong unit, typo) or **real value** (rare but possible, typical for its group)?

## 4. One category: bar of counts

`geom_bar()` counts the rows for you. Use `geom_col()` when the table already holds the values.

```{r}
#| label: bar
penguins |>
  ggplot(aes(x = species)) +
  geom_bar()
```

## 5. A number by a category: box plot

```{r}
#| label: boxplot
penguins |>
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

## 6. More variables without 3D

Each new variable goes into one more channel: color, shape, size, or a facet panel.

```{r}
#| label: more-variables
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = species, shape = sex)) +
  geom_point() +
  facet_wrap(vars(island))
```

## 7. Data quality checklist

Run the five checks of the [data quality checklist](../../capstone/data_quality_checklist.md) on `penguins`: types, missing values, units and ranges, duplicates, row counts.

```{r}
#| label: quality
# Units and ranges: do the minimum and maximum make sense?
penguins |>
  summary()
```

```{r}
#| label: duplicates
# Duplicates: count rows that repeat an earlier row
penguins |>
  summarize(duplicated_rows = sum(duplicated(pick(everything()))))
```

## 8. Your turn

Draw five graphs, one per row of the chart-by-type map, and fill in the map.

| Variables | Chart | Your example |
|---|---|---|
| 1 number | histogram | |
| 2 numbers | scatter | |
| 1 category | bar of counts | |
| number by category | box plot | |
| time | line (next classes) | |

> Deliverable: the chart-by-type map filled with five graphs.
