---
title: "Class 1. Setup check"
format: html
---

Work through the chunks in order. Run each chunk with Ctrl+Shift+Enter (Cmd+Shift+Enter on Mac).

## 1. The console

```{r}
#| label: console
# R is a calculator
1 + 1
```

## 2. Load packages

```{r}
#| label: packages
# pacman installs a package if it is missing, then loads it
pacman::p_load(tidyverse, palmerpenguins, here)
```

## 3. Find your project folder

```{r}
#| label: here
# This path must end with the name of the course folder
here()
```

If the path points to another folder, close RStudio, double-click `data_visualization_course.Rproj`, and run this chunk again.

## 4. Read a file with a relative path

```{r}
#| label: read
here("data", "life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  glimpse()
```

## 5. Draw one graph

```{r}
#| label: graph
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

## Check

- [ ] Chunk 1 printed `[1] 2`.
- [ ] Chunk 3 printed the project folder.
- [ ] Chunk 4 printed a table with `country` and year columns.
- [ ] Chunk 5 drew a scatter plot.

Fix any red text with the teacher before you leave.
