---
title: "Class 2. Grammar of graphics: data, aesthetics, geometry"
format: html
---

**Question for today:** Do penguins with longer flippers weigh more?

## Setup

```{r}
#| label: setup
# One line loads every package we need. pacman installs missing ones.
pacman::p_load(tidyverse, palmerpenguins, here)
```

## 1. Look at the data

```{r}
#| label: look
# The pipe |> sends the data on its left into the function on its right
penguins |>
  glimpse()
```

## 2. Data + aes + geom

Every ggplot has three parts: the **data**, the **aesthetics** (which variable goes where), and the **geometry** (which shape draws it).

```{r}
#| label: first-scatter
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

## 3. Add a third variable with color

```{r}
#| label: color
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

## 4. Your turn

Change `x` and `y` to bill length and bill depth. Write one sentence about what you see.

```{r}
#| label: your-turn
#| eval: false
penguins |>
  ggplot(aes(x = ___, y = ___, color = species)) +
  geom_point()
```

## 5. Save the graph in the project

```{r}
#| label: save
#| eval: false
# here() builds a path from the project folder, so it works on any computer
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()

# ggsave() saves the last plot you drew
ggsave(here("class_02_first_graph.png"), width = 6, height = 4)
```

> Deliverable: one graph, one sentence that answers the question, saved in your project folder.
