---
title: "Class 2. Setup check"
format: html
---

Run each chunk with Ctrl+Shift+Enter (Cmd+Shift+Enter on Mac). Work in order.

## 0. Is the project open?

Look at the top right corner of RStudio. It must show `data_visualization_course`. If it shows `Project: (None)`, close RStudio and double-click `data_visualization_course.Rproj`.

## 1. The console

```{r}
#| label: console
# R is a calculator
1 + 1
```

## 2. Install pacman (run once)

```{r}
#| label: install-pacman
#| eval: false
install.packages("pacman")
```

## 3. Install and load the packages (run once; it takes several minutes)

`p_load()` installs a package if it is missing, then loads it.

```{r}
#| label: install-packages
#| eval: false
pacman::p_load(tidyverse, here, janitor, gtsummary, palmerpenguins,
               gapminder, NHANES, ggrepel, knitr, rmarkdown)
```

## 4. Load the packages for today

```{r}
#| label: packages
pacman::p_load(tidyverse, here, palmerpenguins)
```

## 5. Find your project folder

```{r}
#| label: here
# This path must end with the name of the course folder
here()
```

## 6. Read a file with a relative path

```{r}
#| label: read
here("data", "life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  glimpse()
```

## 7. Draw one graph

```{r}
#| label: graph
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

## Check

- [ ] The project name shows at the top right of RStudio.
- [ ] Chunk 1 printed `[1] 2`.
- [ ] Chunk 5 printed the project folder.
- [ ] Chunk 6 printed a table with `country` and year columns.
- [ ] Chunk 7 drew a scatter plot.

Fix any red text with the teacher before you leave.
