---
title: "Class 4. Row and column verbs"
format: html
---

**Questions for today:** Three questions about adults in the NHANES survey. You answer each with one verb and one plot.

```{r}
#| label: setup
pacman::p_load(tidyverse, NHANES, here)
```

## 1. filter(): choose rows

```{r}
#| label: filter
# Q1. Do adults who sleep less have higher BMI?
NHANES |>
  filter(Age >= 18, !is.na(SleepHrsNight), !is.na(BMI)) |>
  ggplot(aes(x = SleepHrsNight, y = BMI)) +
  geom_jitter(alpha = 0.2)
```

Logical operators: `==` equal, `!=` not equal, `>` `<` `>=` `<=`, `&` and, `|` or, `%in%` in a set, `!is.na()` not missing.

## 2. select(): choose columns

```{r}
#| label: select
NHANES |>
  select(Age, Gender, BMI, SleepHrsNight, PhysActive) |>
  glimpse()
```

## 3. mutate() and case_when(): make new columns

```{r}
#| label: mutate
# Q2. How many adults fall in each BMI group?
NHANES |>
  filter(Age >= 18, !is.na(BMI)) |>
  mutate(bmi_group = case_when(
    BMI < 18.5 ~ "Underweight",
    BMI < 25   ~ "Normal",
    BMI < 30   ~ "Overweight",
    .default   = "Obese"
  )) |>
  ggplot(aes(x = bmi_group)) +
  geom_bar()
```

## 4. arrange(): sort rows

```{r}
#| label: arrange
# Q3. Which 10 adults sleep the fewest hours?
NHANES |>
  filter(Age >= 18, !is.na(SleepHrsNight)) |>
  arrange(SleepHrsNight) |>
  select(Age, Gender, SleepHrsNight, BMI) |>
  slice_head(n = 10)
```

## 5. Your turn

Write one question about NHANES. Answer it with at least one `filter()`, one `mutate()`, and one plot.

```{r}
#| label: your-turn
#| eval: false
NHANES |>
  filter(___) |>
  mutate(___) |>
  ggplot(aes(___)) +
  geom____()
```

> Deliverable: three questions answered, each with a verb and a plot.
