---
title: "Class 4. Row and column verbs: filter() and select()"
format: html
---

**Questions for today:** Three questions about adults in the NHANES survey. The plot code is written. You write only the condition inside `filter()`.

```{r}
#| label: setup
pacman::p_load(tidyverse, NHANES, here)
```

## Comparisons you will use

| Symbol | Meaning |
|---|---|
| `==` | equal to |
| `!=` | not equal to |
| `>`, `<`, `>=`, `<=` | greater, smaller |
| `%in%` | is one of a set |
| `!is.na()` | is not missing |
| `&` | and (you can also separate conditions with a comma) |

## 1. filter(): choose rows (demo)

```{r}
#| label: filter-demo
# Q1. Do adults who sleep less have a higher BMI?
NHANES |>
  filter(Age >= 18, !is.na(SleepHrsNight), !is.na(BMI)) |>
  ggplot(aes(x = SleepHrsNight, y = BMI)) +
  geom_jitter(alpha = 0.2)
```

## 2. Your turn: complete the condition

```{r}
#| label: filter-2
#| eval: false
# Q2. What is the BMI distribution of women aged 18 to 40?
# Fill the blanks: Gender is "female", Age is between 18 and 40, BMI is not missing
NHANES |>
  filter(___) |>
  ggplot(aes(x = BMI)) +
  geom_histogram(binwidth = 2)
```

```{r}
#| label: filter-3
#| eval: false
# Q3. Does sleep differ between physically active and inactive adults?
# Fill the blank: adults (18 or older) with known PhysActive and SleepHrsNight
NHANES |>
  filter(___) |>
  ggplot(aes(x = PhysActive, y = SleepHrsNight)) +
  geom_boxplot()
```

## 3. select(): choose columns

```{r}
#| label: select
NHANES |>
  select(Age, Gender, BMI, SleepHrsNight, PhysActive) |>
  glimpse()
```

## 4. mutate(): one new column with arithmetic

```{r}
#| label: mutate
# Convert age in years to age in decades
NHANES |>
  filter(Age >= 18, !is.na(BMI)) |>
  mutate(age_decade = Age / 10) |>
  ggplot(aes(x = age_decade, y = BMI)) +
  geom_jitter(alpha = 0.1)
```

## 5. Your turn

Write one question about NHANES. Use `filter()` with at least two conditions and one plot.

> Deliverable: three questions answered with `filter()` and a plot.
