---
title: "Class 8. Principle: Tell"
format: html
---

**Rule 3, Tell:** the title states the finding. Annotations point to the evidence.

```{r}
#| label: setup
pacman::p_load(tidyverse, gapminder, here)
```

## 1. Topic title versus active title

| Topic title | Active title |
|---|---|
| Life expectancy in Poland | Polish life expectancy rose 14 years between 1952 and 2007 |

## 2. labs(): title, subtitle, axes, source

```{r}
#| label: labs
gapminder |>
  filter(country == "Poland") |>
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "Polish life expectancy rose 14 years between 1952 and 2007",
    subtitle = "Years of life at birth",
    x = NULL,
    y = NULL,
    caption = "Source: Gapminder"
  ) +
  theme_minimal()
```

## 3. annotate(): point to the evidence

```{r}
#| label: annotate
gapminder |>
  filter(country == "Poland") |>
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(color = "firebrick", linewidth = 1) +
  annotate("text", x = 1952, y = 63, label = "61 years in 1952",
           hjust = 0, color = "gray30") +
  annotate("text", x = 2007, y = 74, label = "76 years in 2007",
           hjust = 1, color = "gray30") +
  labs(
    title = "Polish life expectancy rose 14 years between 1952 and 2007",
    subtitle = "Years of life at birth",
    x = NULL, y = NULL,
    caption = "Source: Gapminder"
  ) +
  theme_minimal()
```

## 4. Your turn

Take your before and after graph from class 7. Add an active title, a source, and one annotation.

> Deliverable: one graph with an active title and one annotation.
