---
title: "Exercise. Add the continent to each country"
format: html
---

**Question:** Which continent does each country belong to, and does the life expectancy of the continents differ in 2023?

The Gapminder table in `data/gapminder_life_expectancy_wide.csv` has one row per country and one column per year (1800 to 2100). It has no continent. The package `countrycode` adds it.

```{r}
#| label: setup
# Install once with: install.packages("pacman")
pacman::p_load(tidyverse, here, janitor, countrycode)
```

## 1. Look at the table

```{r}
#| label: look
# geo = ISO3 code in lower case, name = country name, then one column per year
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(geo, name, `1800`, `2023`, `2100`) |>
  glimpse()
```

Years after 2023 are projections. Use 1800 to 2023 as data.

## 2. Add the continent from the country name

`countrycode()` takes the vector, the format you have (`origin`), and the format you want (`destination`).

```{r}
#| label: continent-from-name
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(geo, name, `2023`) |>
  mutate(continent = countrycode(name, origin = "country.name", destination = "continent")) |>
  glimpse()
```

## 3. Check for countries without a continent

A name that `countrycode` does not recognise returns `NA` and a warning. Check it every time.

```{r}
#| label: check-name-matches
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(geo, name) |>
  mutate(continent = countrycode(name, origin = "country.name", destination = "continent")) |>
  filter(is.na(continent))
```

Write down how many countries have no continent: [YOUR ANSWER HERE]

## 4. Use the ISO3 code instead

The column `geo` holds ISO3 codes (`usa`, `lva`, `chl`). Codes are less ambiguous than names such as "UK", "Lao" or "Congo, Dem. Rep.". Compare both routes.

```{r}
#| label: continent-from-iso
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(geo, name) |>
  mutate(
    continent_name = countrycode(name, origin = "country.name", destination = "continent"),
    continent_iso  = countrycode(str_to_upper(geo), origin = "iso3c", destination = "continent")
  ) |>
  # Rows where the two routes disagree, or one is missing
  filter(is.na(continent_name) | is.na(continent_iso) | continent_name != continent_iso)
```

Which route do you trust more, and why? [YOUR ANSWER HERE]

## 5. Fix what is left

If a country is still missing, add it by hand with `case_when()`.

```{r}
#| label: fix-missing
#| eval: false
# Replace ___ with the code of a country from step 4, and with its continent
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  select(geo, name) |>
  mutate(
    continent = countrycode(str_to_upper(geo), origin = "iso3c", destination = "continent"),
    continent = case_when(
      is.na(continent) & geo == "___" ~ "___",
      .default = continent
    )
  ) |>
  filter(is.na(continent))
```

## 6. Count countries per continent

```{r}
#| label: count-continent
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(continent = countrycode(str_to_upper(geo), origin = "iso3c", destination = "continent")) |>
  count(continent, sort = TRUE) |>
  ggplot(aes(x = n, y = fct_reorder(continent, n))) +
  geom_col() +
  labs(x = "Countries", y = NULL, title = "Countries per continent in the Gapminder table")
```

## 7. Reshape and compare continents

```{r}
#| label: continents-2023
here("data", "gapminder_life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  mutate(continent = countrycode(str_to_upper(geo), origin = "iso3c", destination = "continent")) |>
  pivot_longer(
    cols = `1800`:`2100`,
    names_to = "year",
    values_to = "life_exp",
    names_transform = as.integer
  ) |>
  filter(year == 2023, !is.na(life_exp), !is.na(continent)) |>
  ggplot(aes(x = life_exp, y = fct_reorder(continent, life_exp, .fun = median))) +
  geom_boxplot() +
  labs(x = "Life expectancy in 2023 (years)", y = NULL)
```

## Your turn

Join `data/rosling_tidy.csv` (continent from the gapminder package) to your continents by country name and list the countries where the two continents differ: [YOUR ANSWER HERE]
