---
title: "Class 5. Tidy data: from wide to long"
format: html
---

**Question for today:** How did life expectancy change in five countries since 1952?

```{r}
#| label: setup
pacman::p_load(tidyverse, here)
```

## Tidy data rules

1. Each variable is a column.
2. Each observation is a row.
3. Each value is a cell.

## 1. Read the wide table

```{r}
#| label: read-wide
# Years are column names. This table is wide.
here("data", "life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  glimpse()
```

## 2. Try to plot it

We want `year` on the x axis, but `year` is not a column. ggplot cannot draw this table as it is.

## 3. pivot_longer(): wide to long

```{r}
#| label: pivot
here("data", "life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  pivot_longer(
    cols = `1952`:`2007`,       # the year columns
    names_to = "year",          # old column names go here
    values_to = "life_exp",     # old cell values go here
    names_transform = as.integer
  ) |>
  glimpse()
```

## 4. Time variable: line plot

```{r}
#| label: line
here("data", "life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  pivot_longer(`1952`:`2007`, names_to = "year", values_to = "life_exp",
               names_transform = as.integer) |>
  filter(country %in% c("Finland", "Sweden", "Poland", "Japan", "Brazil")) |>
  ggplot(aes(x = year, y = life_exp, color = country)) +
  geom_line()
```

## 5. Your turn

Pick five other countries. Describe the largest change in one sentence.

> Deliverable: a reshaped table and a line plot of five countries.
