---
title: "Capstone: [your title with the finding]"
author: "[your name] (and partner, if you work in pairs)"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup
pacman::p_load(tidyverse, here, janitor)
```

## M1. Ask (class 15)

**Question:** [one question that names a number, a group, and a period]

**Expected message:** [one sentence]

**Source:** [dataset name and link]

```{r}
#| label: import
#| eval: false
# Replace the path with your file in the data/ folder
here("data", "my_data.csv") |>
  read_csv(show_col_types = FALSE) |>
  clean_names() |>
  glimpse()
```

Data quality: run the five checks of the [data quality checklist](data_quality_checklist.md) and write one line per check.

Tidy check (answer yes or no):

- Each variable is a column.
- Each observation is a row.
- Each value is a cell.

## M2. Wrangle (class 16)

```{r}
#| label: wrangle
#| eval: false
# One pipeline: filter, mutate, pivot, join, group_by + summarize
here("data", "my_data.csv") |>
  read_csv(show_col_types = FALSE) |>
  clean_names() |>
  filter() |>
  mutate() |>
  group_by() |>
  summarize()
```

## M3. Draft (class 17)

Chart type, chosen from the data types (see class 3):

| Variables | Chart |
|---|---|
| 1 number | histogram |
| 2 numbers | scatter |
| 1 category | bar of counts |
| number by category | box plot |
| time | line |

```{r}
#| label: draft
#| eval: false
# Continue the wrangle pipeline and pipe it into ggplot
here("data", "my_data.csv") |>
  read_csv(show_col_types = FALSE) |>
  ggplot(aes()) +
  geom_point()
```

## M4. Refine (class 18)

Checklist:

- [ ] Ask: the title answers the question.
- [ ] Show: right chart, bars start at zero, scales are honest.
- [ ] Cut: no border, no redundant legend, no extra grid lines.
- [ ] Focus: gray plus one color.
- [ ] Tell: active title, source, one annotation.

```{r}
#| label: final
#| eval: false
# Paste the final plot code here, then save the last plot
ggsave(here("capstone_final.png"), width = 8, height = 5.5, dpi = 300)
```

**60-second script:** [beginning, middle, end]
