---
title: "Class 11. Small multiples: compare 1952 and 2007"
format: html
---

**Question for today:** How did the world change between 1952 and 2007?

```{r}
#| label: setup
pacman::p_load(tidyverse, here)
```

## 1. facet_wrap(): one panel per year

```{r}
#| label: facet-wrap
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year %in% c(1952, 2007)) |>
  ggplot(aes(x = income, y = life_exp, size = population, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_size_area(max_size = 12) +
  facet_wrap(vars(year))
```

## 2. facet_grid(): rows by one variable, columns by another

```{r}
#| label: facet-grid
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year %in% c(1952, 2007)) |>
  ggplot(aes(x = income, y = life_exp, size = population)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_size_area(max_size = 10) +
  facet_grid(rows = vars(continent), cols = vars(year))
```

## 3. Keep scales equal across panels

By default all panels share the same axes, which is what you want when you compare. Do not use `scales = "free"` unless you explain why.

## 4. Your turn

Add a third year (1980 is not in the data; use 1977 or 1982). Which panel arrangement shows the change best, and why?

> Deliverable: the faceted Rosling chart for 1952 and 2007.
