---
title: "Class 9. Joins: build the Rosling table"
format: html
---

**Question for today:** Can we build one tidy table from three wide tables?

```{r}
#| label: setup
pacman::p_load(tidyverse, here)
```

## 1. Three wide tables

```{r}
#| label: look
# Each table has one row per country and one column per year
here("data", "income_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  glimpse()
```

## 2. Keys

A key is the column (or columns) that identify a row. After reshaping, the key is `country` + `year`. A join matches rows by key.

## 3. Reshape each table, then join

`left_join()` keeps every row of the table on the left and adds the matching columns from the table on the right.

```{r}
#| label: join
# Reshape the three tables, join them, and save the tidy result
here("data", "life_expectancy_wide.csv") |>
  read_csv(show_col_types = FALSE) |>
  pivot_longer(`1952`:`2007`, names_to = "year", values_to = "life_exp",
               names_transform = as.integer) |>
  left_join(
    here("data", "income_wide.csv") |>
      read_csv(show_col_types = FALSE) |>
      pivot_longer(`1952`:`2007`, names_to = "year", values_to = "income",
                   names_transform = as.integer),
    by = c("country", "year")
  ) |>
  left_join(
    here("data", "population_wide.csv") |>
      read_csv(show_col_types = FALSE) |>
      pivot_longer(`1952`:`2007`, names_to = "year", values_to = "population",
                   names_transform = as.integer),
    by = c("country", "year")
  ) |>
  write_csv(here("data", "rosling_tidy.csv"))
```

## 4. Check the row count

Life expectancy has 142 countries and 12 years, so the table must have 1,704 rows. More rows mean a key repeated. Fewer rows mean a join dropped rows.

```{r}
#| label: check
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  summarize(rows = n(), countries = n_distinct(country), missing = sum(is.na(income)))
```

## 5. First 2007 scatter

```{r}
#| label: scatter-2007
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp)) +
  geom_point()
```

> Deliverable: `data/rosling_tidy.csv` (country, continent, year, life_exp, income, population) and a first 2007 scatter.
