---
title: "Class 6. Split, apply, combine"
format: html
---

**Questions for today:** How did life expectancy change by continent? What is the mean BMI by group in NHANES?

```{r}
#| label: setup
pacman::p_load(tidyverse, gapminder, NHANES, here)
```

## The idea

1. **Split** the table into groups: `group_by()`.
2. **Apply** a function to each group: `summarize()`.
3. **Combine** the results into a new table (dplyr does this for you).

## 1. count(): the shortest summary

```{r}
#| label: count
gapminder |>
  filter(year == 2007) |>
  count(continent)
```

## 2. group_by() + summarize()

```{r}
#| label: sac
# Population-weighted mean life expectancy by continent and year
gapminder |>
  group_by(continent, year) |>
  summarize(life_exp = weighted.mean(lifeExp, pop), .groups = "drop")
```

## 3. Summary table into a line plot

```{r}
#| label: sac-line
gapminder |>
  group_by(continent, year) |>
  summarize(life_exp = weighted.mean(lifeExp, pop), .groups = "drop") |>
  ggplot(aes(x = year, y = life_exp, color = continent)) +
  geom_line()
```

## 4. Summary table into a bar chart

Here the table already holds the values, so we use `geom_col()`.

```{r}
#| label: sac-col
gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarize(life_exp = weighted.mean(lifeExp, pop)) |>
  ggplot(aes(x = continent, y = life_exp)) +
  geom_col()
```

## 5. NHANES: mean BMI by group

```{r}
#| label: nhanes
NHANES |>
  filter(Age >= 18, !is.na(BMI), !is.na(PhysActive)) |>
  group_by(Gender, PhysActive) |>
  summarize(mean_bmi = mean(BMI), n = n(), .groups = "drop") |>
  ggplot(aes(x = PhysActive, y = mean_bmi, fill = Gender)) +
  geom_col(position = "dodge")
```

## 6. Your turn

Compute the median income per person (`gdpPercap`) by continent for 1952 and 2007. Plot it.

> Deliverable: life expectancy by continent over time, and one NHANES mean by group.
