---
title: "Class 7. Principles I: Cut and Focus"
format: html
---

**Rule 1, Cut:** remove every mark that does not carry data.
**Rule 2, Focus:** gray for context, one color for the message.

```{r}
#| label: setup
pacman::p_load(tidyverse, gapminder, here)
```

## 1. Before

```{r}
#| label: before
gapminder |>
  filter(continent == "Europe") |>
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_line()
```

Close your eyes test: look for 5 seconds, close your eyes, say what you remember. Most people remember a mess of colors.

## 2. Cut: remove non-data ink

```{r}
#| label: cut
gapminder |>
  filter(continent == "Europe") |>
  ggplot(aes(x = year, y = lifeExp, group = country)) +
  geom_line(color = "gray70") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

## 3. Focus: one color for the message

```{r}
#| label: focus
gapminder |>
  filter(continent == "Europe") |>
  mutate(highlight = country == "Poland") |>
  ggplot(aes(x = year, y = lifeExp, group = country,
             color = highlight, alpha = highlight)) +
  geom_line() +
  scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "gray70")) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.6)) +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

## 4. Your turn

Take one graph from classes 3 to 6. Make a before and after pair. List what you cut and what you highlighted.

> Deliverable: one before and after pair with a list of changes.
