---
title: "Class 12. Rosling polish and export"
format: html
---

**Goal:** a final 2007 chart that passes the five-principle checklist.

```{r}
#| label: setup
pacman::p_load(tidyverse, ggrepel, here)
```

## 1. Alpha for overlap, direct labels for the story

```{r}
#| label: final
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp, size = population, color = continent)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    # Label only the five most populous countries
    data = \(d) slice_max(d, population, n = 5),
    aes(label = country),
    size = 3.5, color = "gray20", box.padding = 1.5, show.legend = FALSE
  ) +
  scale_x_log10(labels = scales::label_dollar(accuracy = 1)) +
  scale_size_area(max_size = 15, guide = "none") +
  labs(
    title = "In 2007, countries above $10,000 per person lived 15 years longer than the rest",
    subtitle = "Life expectancy at birth, by income per person (log scale). Circle area shows population.",
    x = "Income per person (international dollars)",
    y = "Life expectancy (years)",
    color = NULL,
    caption = "Source: Gapminder"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "top")
```

## 2. Advanced customization: manual colors and theme

Set colors by name so each continent keeps the same color in every graph of your report. Move the title to the left edge of the plot and control text size and spacing with `theme()`.

```{r}
#| label: custom-theme
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp, size = population, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::label_dollar(accuracy = 1)) +
  scale_size_area(max_size = 15, guide = "none") +
  # A named vector fixes one color per continent
  scale_color_manual(values = c(Africa = "#E69F00", Americas = "#009E73",
                                Asia = "#D55E00", Europe = "#0072B2",
                                Oceania = "#CC79A7")) +
  labs(title = "In 2007, countries above $10,000 per person lived 15 years longer than the rest",
       x = "Income per person (log scale)", y = "Life expectancy (years)",
       color = NULL, caption = "Source: Gapminder") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title.position = "plot",                       # align title with the plot edge
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(color = "gray40"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "gray30"),
    legend.position = "top"
  )
```

## 3. Five-principle checklist

| Principle | Check |
|---|---|
| Ask | Does the title answer the question? |
| Show | Is the data tidy, the geom right, the scale honest (log labeled, area for size)? |
| Cut | Did you remove grid lines, borders, and legends that carry nothing? |
| Focus | Does one thing stand out? |
| Tell | Does the title state a finding, and does an annotation point to it? |

## 4. Export at 300 dpi

```{r}
#| label: export
#| eval: false
# ggsave() saves the last plot you drew
ggsave(here("rosling_2007.png"), width = 8, height = 5.5, dpi = 300)
```

> Deliverable: the final Rosling PNG at 300 dpi.
