---
title: "Class 12. Rosling polish and export"
format: html
---

**Goal:** a final 2007 chart that passes the five-principle checklist.

```{r}
#| label: setup
pacman::p_load(tidyverse, ggrepel, here)
```

## 1. Alpha for overlap, direct labels for the story

```{r}
#| label: final
here("data", "rosling_tidy.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(year == 2007) |>
  ggplot(aes(x = income, y = life_exp, size = population, color = continent)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    # Label only the five most populous countries
    data = \(d) slice_max(d, population, n = 5),
    aes(label = country),
    size = 3.5, color = "gray20", box.padding = 1.5, show.legend = FALSE
  ) +
  scale_x_log10(labels = scales::label_dollar(accuracy = 1)) +
  scale_size_area(max_size = 15, guide = "none") +
  labs(
    title = "In 2007, countries above $10,000 per person lived 15 years longer than the rest",
    subtitle = "Life expectancy at birth, by income per person (log scale). Circle area shows population.",
    x = "Income per person (international dollars)",
    y = "Life expectancy (years)",
    color = NULL,
    caption = "Source: Gapminder"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "top")
```

## 2. Five-principle checklist

| Principle | Check |
|---|---|
| Ask | Does the title answer the question? |
| Show | Is the data tidy, the geom right, the scale honest (log labeled, area for size)? |
| Cut | Did you remove grid lines, borders, and legends that carry nothing? |
| Focus | Does one thing stand out? |
| Tell | Does the title state a finding, and does an annotation point to it? |

## 3. Export at 300 dpi

```{r}
#| label: export
#| eval: false
# ggsave() saves the last plot you drew
ggsave(here("rosling_2007.png"), width = 8, height = 5.5, dpi = 300)
```

> Deliverable: the final Rosling PNG at 300 dpi.
